---
title: "2311347_MA321-7-SP_Initial_Task"
output:
  html_document:
    df_print: paged
  pdf_document: default
date: "2024-02-07"
---

```{r eval=TRUE}

library(MASS)
library(kableExtra)

```

```{r eval=TRUE}

knitr::opts_chunk$set(echo = TRUE)

```

### R Markdown

### Read the data into RStudio (or R) using the read.csv R command

```{r eval=TRUE}

InitialData <- read.csv(file="gene-expression-invasive-vs-noninvasive-cancer.csv")

```

### Check using the str, dim and dimnames command it worked - parts of the generated output are added as comments; lines starting with "#" comments and not R code. 

```{r eval=TRUE}

str(InitialData)

```

### 'data.frame':	78 obs. of  4949 variables:
### $ J00129        : num  -0.448 -0.48 -0.568 -0.819 -0.112 -0.391 -0.624 -0.528 -0.811 -0.839 ...
### $ Contig29982_RC: num  -0.296 -0.512 -0.411 -0.267 -0.67 -0.31 -0.12 -0.447 -0.536 2 ...
### ...


```{r eval=TRUE}

dim(InitialData)

```

### [1]   78 4949
 

```{r eval=TRUE}

dimnames(InitialData)[[2]][4947:4949]

```


### [1] "NM_000898" "AF067420"  "Class" 


### The data set has 78 rows (patients) 4949 columns with 4948 gene expression measurement of cancer tissue, each column representing a 'gene' with column 4949 having the information of a class variable with two values: 1 and 2.

```{r eval=TRUE}

table(InitialData[4949])

```

### Class
### 1  2 
### 34 44 

### To gurantee that your data narrative is individual we ask you to set the seed of the random number generator with R function set.seed using your Registration Number; for example "2244222" - replace by your registration number

```{r eval=TRUE}

set.seed(2311347)

```

### Note: R function set.seed generates no R output. Make sure that you run your final data analysis with the submitted R code and that all tables and figures are generated by the final code. You may use R Mark Down or similar R tools to support this.

### Selects your random individual subset of 10 genes
 
```{r eval=TRUE}

my.gene.subset <- rank(runif(1:4948))[1:10] 
my.gene.subset
rank(runif(1:4948))[1:10]

```

###  [1] 4136 4878  175 1409 3208   69 1571 4844  211 3325

### Store sample data of 10 random gene data into a variable. 
```{r eval=TRUE}

my_gene_subset_data <- InitialData[my.gene.subset]

```

### Check whether dataset has any null values or not
```{r eval=TRUE}

sum(is.na(my_gene_subset_data))

```

### 1.a) Compute the variance, co-variance and correlation matrix of your random subset of 10 genes. Add an appropriate table to your report.

### Calculating the variance of the sample random gene data.
```{r eval=TRUE}

my_gene_var <- var(my_gene_subset_data)
my_gene_var
my_gene_var_table <- kable(my_gene_var, digit=3)
my_gene_var_table

```

## Observations and Findings

Variance is used to define how the data is spread. Lower variance values indicates how far it is spread from the mean and vice-versa .From the above result, we can observe that the most of the data falls near the mean. 

The minimum variance value is 0.018 

The maximum variance value is 0.222



### Calculating the co-relation of the sample random gene data.
```{r eval=TRUE}

my_gene_cor <- cor(my_gene_subset_data)
my_gene_cor
my_gene_cor_table <- kable(my_gene_cor, digit=3)
my_gene_cor_table
min(my_gene_cor)
max(my_gene_cor)
```

## Observations and Findings

From the above result, we can observe that the most of the data values are positive which indicates that their is a better positive relation between the data elements than negative relation. Positive relation indicates that elements are directly proportional i'e if one variable value increases other variable value also increases. 

The minimum co-relation value is -0.404527 

The maximum co-relation value is 1. 

We can conclude that there is no elements in data set which doesn't have any relation.



### Calculating the co-variance of the sample random gene data.
```{r eval=TRUE}

my_gene_cov <- cov(my_gene_subset_data)
my_gene_cov
mean(my_gene_cov)
my_gene_cov_table <- kable(my_gene_cov, digit=3)
my_gene_cov_table
min(my_gene_cov)
max(my_gene_cov)

```

## Observations and Findings

From the above result, we can observe that the most of the data values are positive which indicates that their values increases where other values also increases. 

The minimum variance value is -0.0348 

The maximum variance value is 0.2490



### 1.b) Using R to calculate the distance matrix of your random subset of 10 genes. Add an appropriate table to your report.
```{r eval=TRUE}

dist_matrix <- dist(my_gene_subset_data)
as_matrix <- as.matrix(dist_matrix)

kable(as_matrix[,1:28], digit=3)
kable(as_matrix[,29:53], digit=3)
kable(as_matrix[,54:78], digit=3)

```

## Observations and Findings

Distance matrix is a square matrix used to calculate the distance between the elements of the data set. If the distance of any two pair of elements have same distance, we can say that those elements have similarities and vice-versa. 

The minimum distance between two elements in dataset is 0.25002

The maximum distance between two elements in dataset is 3.302722


### 1.c) Using R to calculate univariate Q-Q-plots and a Q-Q-plot based on the generalised distance for the observations of your random subset of 10 genes. Add appropriate figures to your report.
```{r eval=TRUE}
par(mfrow = c(2, 5))
for (i in my.gene.subset){
  qqnorm(InitialData[[i]], main= paste('Q-Q plot for', names(InitialData[i])))
  qqline(InitialData[[i]])
}

cm <- colMeans(my_gene_subset_data)
S <- my_gene_cov
d <- apply(my_gene_subset_data, MARGIN = 1, function(x) 
  t(x - cm) %*% solve(S) %*% (x - cm))
par(mfrow = c(1,1))
plot(qchisq((1:nrow(my_gene_subset_data) - 1/2) / nrow(my_gene_subset_data), df = 9), sort(d),
     xlab = expression(paste(chi[3]^2, " Quantile")), 
     ylab = "Ordered distances"); abline(a = 0, b = 1)

```

## Observations and Findings

univariate Q-Q plot is used to determine whether the data is distributed normally or not. If most of the data points falls on the straight line we can say that data is normally distributed. From the univariate Q-Q plot, we can observe that NM_015710, NM_000274, NM_012081, NM_016618, NM_020188 elements tends to follow normal distribution. 

Generalized Q-Q plot is used to determine the model fitness. If most of the data/residual points falls on the straight line, we can consider that the model is well-fitted. From the above plot, we can observe that points at the right side are away from the straight line indicating that there are some outliers indicating non linearity in data points with response variables.

### 2) Use R for a principal component analysis of your random subset of 10 genes. Add appropriate tables and figures to your report.
```{r eval=TRUE}

my_gene_principle_comp <- princomp(my_gene_subset_data, cor = FALSE)
my_gene_principle_comp
names(my_gene_principle_comp)
summary(my_gene_principle_comp, loadings = TRUE,cutoff=.1)
plot(my_gene_principle_comp)
pairs(my_gene_principle_comp$scores)

```

## Observations and Findings

Principle component analysis ( PCA ) is a dimensionality reduction method used to reduce dimensionality of the datasets by transforming large set of elements into smaller sets. 

From the above calculated results and plots, we can observe that first principle component has 36% of variance on the entire dataset and second principle component has 24% of the variance on the entire dataset. Therefore, Cumulative proportion of first and second principle component explains overall 60% of variance in the entire dataset.

Pairs plot is used to determine the co-relation between pair of classes. From this plots, we can observe the linearity, outliers, clusters, spread of data between classes etc.



### 3.a) Fit a multivariate analysis of variance model (MANOVA) to your random subset of 10 genes. Investigate if there is a difference between invasive (label 1) and noninvasive (label 2) cancer. Note: You need to add column 4949 containing the information invasive and noninvasive cancer to your random subset of 10 genes. Add appropriate tables and figures to your report.
```{r eval=TRUE}

my_gene_subset_data$Class <- InitialData$Class

my_gene_subset_matrix <- as.matrix(my_gene_subset_data[,c(1:10)])

my_gene_subset_manova_fit <- manova(my_gene_subset_matrix~my_gene_subset_data$Class)

summary(my_gene_subset_manova_fit,intercept=TRUE)

```

## Observations and Findings

MANOVA is used determine whether there is any significance differences between groups with dependent variables. If the calculated p-value is less than the significance level i'e p-value < 0.05, we can say that there is a significant difference between groups. From the above observation, we say find that the p-value is 0.0042 which is less that the significance value. So, we can conclude that there is significance difference between invasive and non-invasive cancer. 


### 3.b) Use the first and second principal component to illustrate, if there is a difference between invasive and noninvasive cancer. Add appropriate tables and figures to your report.
```{r eval=TRUE}

my_gene_subset_manova_fit_one_two <- manova(my_gene_principle_comp$scores[,1:2]~my_gene_subset_data$Class)

summary(my_gene_subset_manova_fit_one_two, intercept=TRUE)

```

## Observations and Findings

When we use MANOVA to analyze first and second PCA components, we can observer that the p-value has been increases to 0.07 which is greater than significance value i'e 0.05 from the overall PCA components p-value i'e 0.0042. Hence we can conculude that there is no significant difference between invasive and non-invasive cancer.  

### 4.a) Apply LDA to your random subset of 10 genes and the class variable (invasive (label 1) and noninvasive (label 2) cancer). Calculate a confusion matrix, sensitivity, specificity and misclassification error. Add appropriate tables and figures to your report.
```{r eval=TRUE}

my_gene_subset_lda <- lda(Class~., data=my_gene_subset_data)
my_gene_subset_lda
my_gene_subset_data$Class

my_gene_subset_lda_predicted <- predict(my_gene_subset_lda)$class
my_gene_subset_lda_predicted

table(my_gene_subset_lda_predicted, my_gene_subset_data$Class) 

table(my_gene_subset_data$Class)
table(my_gene_subset_lda_predicted)

misclassification_error_lda <- (table(my_gene_subset_lda_predicted, my_gene_subset_data$Class)[1,2] + 
    table(my_gene_subset_lda_predicted, my_gene_subset_data$Class)[2,1]) / (dim(my_gene_subset_data)[1])
misclassification_error_lda

sensitivity_error_lda <- (table(my_gene_subset_lda_predicted, my_gene_subset_data$Class)[2,2] / 
                        (table(my_gene_subset_lda_predicted, my_gene_subset_data$Class)[2,1]+table(my_gene_subset_lda_predicted, my_gene_subset_data$Class)[2,2]))
sensitivity_error_lda


specificity_error_lda <- (table(my_gene_subset_lda_predicted, my_gene_subset_data$Class)[1,1] / 
                        (table(my_gene_subset_lda_predicted, my_gene_subset_data$Class)[1,1]+table(my_gene_subset_lda_predicted, my_gene_subset_data$Class)[2,1]))
specificity_error_lda

```

## Observations and Findings

LDA is used to understand class distinctions and assess overall classification ability. Additionally included are the confusion matrix, sensitivity, specificity, and misclassification error. These are used to understand the differences between the classes that make up the dependent variable. The confusion matrix is used to interpret the performance of the classification model. Recall, which is a performance statistic also known as sensitivity, is used to evaluate the efficacy model. A performance parameter called specificity—also called precision—is employed to evaluate the effectiveness of a model. Misclassification error is one performance metric used to calculate the model's error rate. The figures presented above demonstrate that the precission is 67%, recall is 76%, and error rate is 25%.



### 4.b) Apply Quadratic discriminant analysis (QDA) to your random subset of 10 genes and the class variable (invasive (label 1) and noninvasive (label 2) cancer). Calculate a confusion matrix, sensitivity, specificity and misclassification error. Add appropriate tables and figures to your report.
```{r eval=TRUE}

my_gene_subset_qda <- qda(Class~., data=my_gene_subset_data)
my_gene_subset_qda
my_gene_subset_data$Class


my_gene_subset_qda_predicted <- predict(my_gene_subset_qda)$class
my_gene_subset_qda_predicted

confusion_matrix <- table(my_gene_subset_qda_predicted, my_gene_subset_data$Class) 
confusion_matrix
table(my_gene_subset_data$Class)
table(my_gene_subset_qda_predicted)

misclassification_error_qda <- (confusion_matrix[2] + confusion_matrix[3]) / (dim(my_gene_subset_data)[1])
misclassification_error_qda

sensitivity_error_qda <- (confusion_matrix[1] / (confusion_matrix[1]+confusion_matrix[2]))
sensitivity_error_qda


specificity_error_qda <- (confusion_matrix[4] / (confusion_matrix[3]+confusion_matrix[4]))
specificity_error_qda


```
## Observations and Findings

QDA is used to understand class distinctions and assess overall classification ability. Additionally included are the confusion matrix, sensitivity, specificity, and misclassification error. These are used to understand the differences between the classes that make up the dependent variable. The confusion matrix is used to interpret the performance of the classification model. Recall, which is a performance statistic also known as sensitivity, is used to evaluate the efficacy model. A performance parameter called specificity—also called precision—is employed to evaluate the effectiveness of a model. Misclassification error is one performance metric used to calculate the model's error rate. The figures presented above demonstrate that the precission is 88%, recall is 79%, and error rate is 15%.


### 4.c) Discuss the difference between LDA and QDA using the results on your random subset of 10 genes and the class variable (invasive (label 1) and noninvasive (label 2) cancer).

There is a difference between the two models after LDA and QDA is performed on the subset. The results are better when QDA is used in place of LDA. We can observe that the Specificity has increased from 67% to 88%. sensitivity increased from 76% to 79% and the misclassification error rate has been decreased from 25% to 15%. Hence we can conclude that for our dataset, QDA performs better than LDA. 


### 5) Use the median of the first principal component of your random subset of 10 genes to predict the class variable (invasive (label 1) and noninvasive (label 2) cancer). Use Fisher’s Exact test and sensitivity, specificity and Youden index. Add appropriate tables and figures to your report.
```{r eval=TRUE}
pca_median <- median(my_gene_principle_comp$scores[,1])
pca_median

my_new_gene_data <- my_gene_subset_data
my_new_gene_data$prid_median <- NA
my_new_gene_data$prid_median <- ifelse(my_gene_principle_comp$scores[,1] < pca_median, 1, 2)
my_new_gene_data
my_new_confusion_matrix <- table(my_new_gene_data$prid_median, my_new_gene_data$Class)
my_new_confusion_matrix

new_sensitivity_error <- (my_new_confusion_matrix[1] / (my_new_confusion_matrix[1]+my_new_confusion_matrix[2]))
new_sensitivity_error


new_specificity_error <- (my_new_confusion_matrix[4] / (my_new_confusion_matrix[3]+my_new_confusion_matrix[4]))

new_specificity_error

fisher_test <- fisher.test(my_new_confusion_matrix)
fisher_test

youden_index <- (new_sensitivity_error + new_specificity_error) -1
youden_index

```

## Observations and Findings

The Fisher's exact test is a statistical method that can be used to find non random co-relations between two category variables.
Using the predict function, the expected values of the class variable are predicted after the initial principal component analysis (PCA) median has been established.

Fisher's test and principal component analysis are comparable in many aspects. however, Fisher's test takes class labels into account while figuring out the discriminant direction. The fisher's extract test shows that the p-value is less than the significance level, rejecting the null hypothesis.

The precission of the model is demonstrated by its specificity. That amounts to almost 59%. The sensitivity represents the recall of the model, which is roughly 61%.
The youden index is a statistic that is used to determine how well a categorization test is working. Its range is from -1 to +1. It evaluates the total discriminatory power of the model while taking specificity and sensitivity into account. The model indicated above has a yoden index of almost 0.2, suggesting that it is a respectable model.